---
title: "CodeBook"
output: html_document
---

# Course Project Definition
1) Merges the training and the test sets to create one data set
2) Extracts only the measurements on the mean and standard deviation for each measurement
3) Uses descriptive activity names to name the activities in the data set
4) Appropriately labels the data set with descriptive variable names
5) From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject

# 1) Merges the training and the test sets to create one data set
## Download Data
The UCI HAR Dataset is downloaded from: [Dataset](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip)

```{r eval = FALSE}
fileurl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
if (!file.exists("./UCI HAR Dataset")) {
  #don't redownload/extract if data already exists
  download.file(fileurl,'./uciDataset.zip', mode = 'wb')
  unzip("uciDataset.zip", exdir = getwd())
}
```

## Import into R for Editing
*featuresList*-- features are read in from the dataset's list and describe the datapoints that were collected <br />
*activityList*-- activities are also read in from the dataset's list and describe what physical activity the subject was doing when the datapoints were collected <br />
*train/testX*-- values of collected datapoints <br />
*train/testY*-- enum of activity being performed during corresponding X's recording <br />
*train/testSubject*-- list of subject's ID numbers who partook in training or testing <br />

```{r eval = FALSE}
featuresList <- read.csv("./UCI HAR Dataset/features.txt",header = FALSE, sep = " ")
featuresList <- as.character(featuresList[,2])
activityList <- read.csv("UCI HAR Dataset/activity_labels.txt", header = FALSE, sep = " ")
activityList <- as.character(activityList[,2])

trainX <- read.table("./UCI HAR Dataset/train/X_train.txt", col.names=featuresList)
trainY <- read.table("./UCI HAR Dataset/train/y_train.txt",header = FALSE, sep = " ",col.names = "Activity")
trainSubject <- read.table("./UCI HAR Dataset/train/subject_train.txt",header = FALSE, sep = " ", col.names="Subject")

testX <- read.table("./UCI HAR Dataset/test/X_test.txt",col.names=featuresList)
testY <- read.table("./UCI HAR Dataset/test/y_test.txt",header = FALSE, sep = " ",col.names = "Activity")
testSubject <- read.table("./UCI HAR Dataset/test/subject_test.txt",header = FALSE, sep = " ", col.names="Subject")
```

# 2) Extracts only the measurements on the mean and standard deviation for each measurement.
## Merge Data Sets
*X*-- Appended *testX* and *trainX* into one table <br />
*Y*-- Appended *trainY* and *trainY* into one table <br />
*Subject*-- Appended *testSubject* and *trainSubject* into one table <br />
*data*-- Merged (by column) the three above tables into one large table. <br />

```{r eval = FALSE}
X <- rbind(testX,trainX)
Y <- rbind(testY,trainY)
Subject <- rbind(testSubject, trainSubject)
data <- cbind(Subject,Y,X)
```
## Extract only Standard Deviation and Mean Columns
*mean_stddev_data*-- Same structure as *data* however columns have been filtered to ones containing only **mean** and **std**
```{r eval=FALSE}
mean_stddev_data <- data %>% select(Subject, Activity, contains("mean"), contains("std"))
```

# 3) Uses descriptive activity names to name the activities in the data set
*tidy_mean_stddev_data*-- Contains the same data as *mean_stddev_data* however has replaced the enum **Activity** column with human readable terms
```{r eval=FALSE}
#replace activity number with english name
tidy_mean_stddev_data <- mean_stddev_data
tidy_mean_stddev_data$Activity <- activityList[tidy_mean_stddev_data$Activity]
```

# 4) Appropriately labels the data set with descriptive variable names
Further modificaiton to *tidy_mean_stddev_data* where columns names are updated from shorthand to human readable camel caped phrases
```{r eval=FALSE}
#make column names human readable, capitilizing for keep camel caps
names(tidy_mean_stddev_data)<-gsub("Acc", "Accelerometer", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("Gyro", "Gyroscope", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("BodyBody", "Body", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("Mag", "Magnitude", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("^t", "Time", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("^f", "Frequency", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("tBody", "TimeBody", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub(".mean()", "Mean", names(tidy_mean_stddev_data), ignore.case = TRUE)
names(tidy_mean_stddev_data)<-gsub(".std()", "STD", names(tidy_mean_stddev_data), ignore.case = TRUE)
names(tidy_mean_stddev_data)<-gsub(".freq()", "Frequency", names(tidy_mean_stddev_data), ignore.case = TRUE)
names(tidy_mean_stddev_data)<-gsub("angle", "Angle", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("gravity", "Gravity", names(tidy_mean_stddev_data))
names(tidy_mean_stddev_data)<-gsub("\\.","",names(tidy_mean_stddev_data))
```

# 5) From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject
*tidy_averages_data* is *tidey_mean_stddev_data* grouped by all **Subject** and **Activity** pairs and the average of each feature within the pair. <br />
*tidy_averages_data* is then written out into a .txt value for evaluation outside of R.

```{r eval=FALSE}
tidy_averages_data <- tidy_mean_stddev_data
tidy_averages_data <- tidy_averages_data %>% group_by(Subject, Activity) %>% summarise_all(mean)

#export this final data set
write.table(tidy_averages_data,file="Averages of Subject and Activity.txt", row.names = FALSE) # row.names removes row numbers so no index column needs to be added
```